Modal Abstractions for Virtualizing Memory Addresses

### 2024-06-30 examples.tex
+

\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\newcommand{\qfraczero}{\textsf{qfrac}}
\newcommand{\true}{\textsf{true}}
\tikzstyle{boxedassert_border} = [sharp corners,line width=0.2pt]
\NewDocumentCommand \boxedassertpv {O{} m o}{%
	\tikz[baseline=(m.base)]{
		%	  \node[rectangle, draw,inner sep=0.8pt,anchor=base,#1] (m) {${#2}\mathstrut$};
		\node[rectangle,inner sep=1.5pt,outer sep=0.2pt,anchor=base] (m) {${\,#2\,}\mathstrut$};
		\draw[#1,boxedassert_border] ($(m.south west)$) rectangle ($(m.north east)$);
	}\IfNoValueF{#3}{^{\,#3}}%
}
\newcommand*{\knowInvpv}[2]{\boxedassertpv{#2}[#1]}
\newcommand*{\ownGhostpv}[2]{\boxedassertpv[dash dot]{#2}[#1]}

@@

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, OFF ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
\subsubsection{Traversing Page Tables}
As we explain in Section \ref{}, a traversal over a page-table tree can be done through the indices into the tables at each level followed by obtaining the address of the next table. In Figure \ref{fig:getnexttable}, we see how to obtain the physicall address of the next level page-table. Traversal start from the root address of the L4 table (\textsf{pte\_t *pml4} as a parameter to \textsf{walkpgdir}). Then, we obtain the address of the entry (\textsf{pml4[PML4EX(va)]}) via the L4 index in the virtual address (\textsf{va}). Finally, we call \textsf{get\_next\_table} to get the virtual address to the next level page-table -- L3.  
\begin{figure*}[ht]
\begin{tabular}{p{0.47\textwidth}p{0.47\textwidth}}
  \begin{lstlisting}[mathescape,basicstyle=\scriptsize\ttfamily]
;;void pte_initialize(pte_t **table) {
...

>>>>

As we explain in Section \ref{}, a traversal over a page-table tree can be done through the indices into the tables at each level followed by obtaining the address of the next table. In Figure \ref{fig:getnexttable}, we see how to obtain the physicall address of the next level page-table. Traversal start from the root address of the L4 table (\textsf{pte\_t *pml4} given as a parameter to \textsf{walkpgdir}). Then, we obtain the address of the entry in the L4 table via the index in the virtual address (\textsf{va}).

Finally, we call \textsf{get\_next\_table} to get the virtual address to the next level page-table -- L3.
Within \textsf{get\_next\_table} first checks if the entry whose address is passed as a parameter -- e.g., for L4 \textsf{pml4[PML4EX(va)]}.
In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (\textsf{kalloc} -- currently an axiom as it is not part of \textsf{pte} module) to be backing the physical frame address of the entry.

Once we make sure that the frame of the page-table entry is backed with an allocated physical page, we can obtain the virtual address of the next level-page table (\textsf{next} of function \textsf{get\_next\_table} in Figure \ref{}) via looking up the \textit{identity mapping} from the physical address (the address value referenced by \textsf{entry->pfn}) to virtual address value to be returned by \textsf{get\_next\_table}.
\subsubsection{Identity Mappings}
Most kernels maintain an invariant on their page tables that the virtual address of any page used for a page table lives at a virtual address
whose value is a constant offset from the physical address a practice sometimes referred to as \emph{identity mapping} (even though the physical-to-virtual translation
is typically not literally the identity function, but adding a constant offset).\footnote{Some kernels do this for all physical memory on the machine, simplifying interaction
with DMA devices.}
\newcommand{\qfraczero}{\textsf{qfrac}}
\newcommand{\true}{\textsf{true}}
\tikzstyle{boxedassert_border} = [sharp corners,line width=0.2pt]
\NewDocumentCommand \boxedassertpv {O{} m o}{%
	\tikz[baseline=(m.base)]{
		%	  \node[rectangle, draw,inner sep=0.8pt,anchor=base,#1] (m) {${#2}\mathstrut$};
		\node[rectangle,inner sep=1.5pt,outer sep=0.2pt,anchor=base] (m) {${\,#2\,}\mathstrut$};
		\draw[#1,boxedassert_border] ($(m.south west)$) rectangle ($(m.north east)$);
	}\IfNoValueF{#3}{^{\,#3}}%
}
\newcommand*{\knowInvpv}[2]{\boxedassertpv{#2}[#1]}
\newcommand*{\ownGhostpv}[2]{\boxedassertpv[dash dot]{#2}[#1]}

@@

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
\[
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner\urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$ 
\mathsf{pte\_initialize}(\mathsf{next})
...
  $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, OFF ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Converting a physical address, whose exsitence is made certain, of a PTE to a virtual address (w/o instruction pointer or flag updates). $R$ abbreviates the resources
associated with \paddr\ by $\Xi$ in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.}
\label{fig:p2v}
\end{figure}

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.

\begin{figure*}[ht]
\begin{tabular}{p{0.47\textwidth}p{0.47\textwidth}}
  \begin{lstlisting}[mathescape,basicstyle=\scriptsize\ttfamily]
;;void pte_initialize(pte_t **table) {
...
